---
title: "Compute Avg Monthly Wage"
author: "Ou Yang Yu"
date: "2024-09-18"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE)
```

# Introduction

This is a job position assessment assigned to me by a Singapore public agency.

I am conducting data analysis using 2 datasets containing information about some companies and its employees.

I have to compute the average monthly wage paid by companies of different sizes.

There were no actual csv files provided.

The Companies dataset have these variables: Company_Name, Date of Incorporation, Revenue, Sector.

The Employees dataset have these variables: Employee_Name, Company_Name, Monthly_Wage

The code below is similar to a pseudocode.

## Load Libraries

```{r libraries}
library(tidyverse)

```

## Read Company and Employees csv files

```{r read-csv}
# Read the CSV files for companies and employees datasets
companies <- read.csv("companies.csv")
employees <- read.csv("employees.csv")

```

## View Dataset

```{r view-dataset}
# Inspect the first few rows of each dataset to ensure they are loaded correctly
head(companies)
head(employees)

```

## Define Company Sizes

```{r define-size}
# Define firm size based on revenue thresholds
# Let's assume small companies have revenue < $1,500,000, medium < $2,000,000, and large >= $2,000,000
companies <- companies %>%
  mutate(firm_Size = case_when(
    Revenue < 1500000 ~ "Small",
    Revenue >= 1500000 & Revenue < 2000000 ~ "Medium",
    Revenue >= 2000000 ~ "Large"
  ))

```

## Merge both Datasets

```{r merge-datasets}
# Merge the two datasets based on Company_Name
merged_data <- employees %>%
  left_join(companies, by = "Company_Name")

```

## Calculate Average Monthly Wage

```{r calc-wage}
# Calculate the average monthly wage for each company size
average_wage_by_firm_size <- merged_data %>%
  group_by(firm_Size) %>%
  summarise(Average_Monthly_Wage = mean(Monthly_Wage, na.rm = TRUE))

```

## Display the Average Monthly Wage Result

```{r display-results}
# Display results
print(average_wage_by_firm_size)

```
